---
title: "Quarto: ojs_define & transpose"
format: html
editor: visual
---

In this document, we show how it is possible to pass a variable from R to Observable. We deal with 3 cases:

-   a simple variable
-   a data frame
-   a spatial data frame (sf)

# 1 - a simple variable

#### {r}

Here we have a variable myvar which is 12.

```{r}
myvar = 12
```

The *ojs_define* instruction simply makes this variable accessible in Observable

```{r}
ojs_define(myvar)
```

#### {ojs}

```{ojs}
myvar
```

# 1 - a dataframe

### {r}

In this second case, we import a csv file.

```{r}
data <- read.csv("data/stat.csv")
```

As before, we use *ojs_define* to make this variable accessible in Observable. We rename this variable to newdata.

```{r}
ojs_define(newdata = data)
```

### {ojs}

Warning, if we visualize this data table in Observable directly, we realize that it does not really have the expected form.

```{ojs}
newdata
```

Thankfully, the *transpose* instruction allows you to convert the table to the correct format.

```{ojs}
data = transpose(newdata)
data
Inputs.table(data)
```

## 3 - a spatial data frame (sf)

### {r}

In this third case, we import a geopakage thanks to the library sf, containing the countries of the world.

```{r, message= FALSE}
library("sf")
countries <- st_read("data/countries.gpkg", quiet = T)
plot(st_geometry(countries))
```

To create a map in Observable, we first need to convert this data set to geojson format. To do this, we use the *geojsonsf* library. Then, the ojs_define() instruction allows to define this variable within ojs cells.

```{r}
library("geojsonsf")
ojs_define(geo = sf_geojson(countries))
```

### {ojs}

Note that with ojs_define, we have passed the variable geo as a string and not actually as an object.

```{ojs}
geo.substr(1, 300)
```

The first thing to do here is to transform our string into a real object. To do this, we use the javascript statement *JSON.parse*.

```{ojs}
countries = JSON.parse(geo) 
countries
```

Now everything is ok to make a map in Observable. See more [here](https://observablehq.com/@neocartocnrs/hello-bertin-js?collection=@neocartocnrs/bertin).

```{ojs}
bertin = import('https://cdn.skypack.dev/bertin@0.9.12')
bertin.draw({
  layers: [
    {
      geojson: countries,
      tooltip:["$ISO3","$NAMEen"]
    }
    ]
})
```

# 4 - Another solution

Another way to pass variables and data from R to Observable is to save and then import data.

### {r}

As before, we import a geopackage in R with *sf*.

```{r}
countries <- st_read("data/countries.gpkg", quiet = T)
```

Then we convert it to geojson and save it in the data directory.

```{r}
write(sf_geojson(countries), "data/world.geojson")
```

### {ojs}

In observable, we just have to import this data file in geojson (or other) format.

```{ojs}
world = FileAttachment("data/world.geojson").json()
```

Let's take a look at it!

```{ojs}
bertin.quickdraw(world)
```

### That's all folks!
